What we {\sf eval} in the shadows

### 2021-04-01 paper/main.tex
3.3

\noindent The pipeline steps are:
\begin{compactenum}
  \item \emph{Download.} The first step consists of downloading and installing
    CRAN packages and acquiring Kaggle corpus. We try to install all available
    CRAN packages. This is somewhat complicated by the fact that many of
    packages depend on native libraries, but these dependencies are not
    explicitly stated in the package description and thus it is hard to
    automate. For Kaggle we use a web crawler to get the programs and the a
    Kaggle official command line tool to download the datasets.
  \item \emph{Runnable code extraction.} Next step is to extract runnabale code.
    For CRAN package we use \genthat~\cite{issta18}, a tool which converts
    runnable code snippets from package documentation and vignettes as well as R
    tests into standalone executable scripts. For Kaggle programs that are not
    already R scripts, we use \c{rmarkdown} and \c{knitr} R packages to extract
    the runnable code from the Rmarkdown or the Jupyter notebook formats
    respectively. Duplicated Kaggle programs are removed. Each of the extracted
    script is wrapped into a call to our dynamic analyzer so we only record eval
    usage triggered by running th script. Without this, the data would include
    eval calls from the unit testing frameworks as well as from bootstrapping R
    virtual machine itself.
  \item \emph{Tracing.} With the scripts ready, we run our dynamic analysis that
    traces the calls to eval and captures other runtime information. For the
    CRAN packages we run two configurations:
  \begin{inparaenum}[(1)]
    \item all of the \CranRunnableScriptsRnd scripts while tracing calls to eval
      originating in any of the \CranAllPackages installed CRAN package, and
    \item a random selection of 1/4 of the scripts while tracing only eval calls
      originating in the core R packages that come bundled with the language.
  \end{inparaenum}
    The reason for this split is that the core R packages are intertwined with
    the language implementation and the usage is different form the user written
    packages hosted on CRAN. For Kaggle, we only trace the eval calls that
    originate in the programs themselves. To avoid any interference, each
    program is run in isolation, in its own R instance.
  \item \emph{Analysis.} In the final step, the output from the dynamic analysis
    is merged, cleaned and summarized in the post-processing phase by custom R
    scripts. The summarized data is then analyzed in RMarkdown notebooks to
    gather insights. Apart from the figures, the data points included in the
    paper are also generated by RMarkdown notebooks as \LaTeX macros.
\end{compactenum}

@@

For this study we need two things: information about calls to eval and related
functions and low-level data such as object allocation and deallocation and
side-effects that happen during eval. The former we can get by injecting code to
the corresponding functions at runtime. For the latter we use
\rdyntrace~\cite{oopsla19a}. It is a modified R virtual machine based on GNU R
4.0.2 that exposes a number of hooks allowing us to get low-level information
about the R interpreter. \rdyntrace only works with the R interpreter and not
the byte-code compiler. However, this does not affect the result since the
compiler does not optimize eval calls.

We use source location to uniquely identify eval call sites. R only retains
source references at the top-level block expressions. A nested call such as
\c{try(eval(x))} does not have a source reference for the \c{eval(x)}. To work
around this, we impute source locations for eval call sites at runtime, by
traversing functions' AST and attaching source reference attribute manually.

The dynamic analysis is implemented as a R package in 1.5K lines of R code and
2K lines of C++.

\mypara{Limitations.} We do not get source locations for \CranUndefinedRnd \eval
calls. In these cases \eval is either passed as an argument to a higher-order
functions or is defined in a function returned by a higher-order function, or it
can be a generated called from package native code. However, this is a meager
\CranUndefinedRatio of all \eval calls and is unlikely to affect our analysis.
We ignore calls to the native \eval function exposed by R. We also ignore the
\c{rlang::tidy\_eval} function which uses native \eval internally because
\c{rlang} is used to implement a DSL for data analysis in R. It introduces a new
first-class promise object called \c{quosure} for which it implements special
evaluation support in \c{tidy\_eval}.

>>>>

\noindent The pipeline steps are:
\medskip
\begin{compactenum}
\item \emph{Download.} Packages are downloaded automatically from CRAN. For
  scripts, a web crawler retrieves code and the Kaggle command line tool
  downloads data. Installation is complicated by native dependencies which
  are not properly documented and thus hard to automatically resolve.
\item \emph{Extract.} Given installed packages and scripts, the next step is to
  create runnable programs. The \genthat tool helpfully extracts all runnable
  code snippets from a package and turns each of these into a self-standing
  program~\cite{issta18}. Some Kaggle kernels are already scripts, for those
  nothing more needs be done. Others kernels are packages as notebooks, either
  as Rmarkdown or Jupyter, for those we use \c{knitr} to extract runnable code.
  The body of each extracted program is instrumented with calls our dynamic
  analyzer to ensure that we only record calls to \eval from the code of
  interest and not from bootstrapping or execution harness operations.
\item \emph{Trace.} Each program is executed using our dynamic analysis tool
  which is heavily instrumented interpreter that captures calls to \eval and
  many other fine-grained runtime events. Packages are run twice, once to
  capture \eval calls originating from package code, and a second time to
  capture calls coming from the base libraries. To avoid any interference, each
  program is run in its own process.
\item \emph{Analyze.} Finally, analysis output is merged, cleaned and summarized
  in a post-processing phase driven by series of R scripts. The summarized data
  is then analyzed in RMarkdown notebooks to gather insights. All figures and
  numbers appearing in the paper are generated automatically. Figures are
  produced in PDF by \c{ggplot2}, numbers are exported as \LaTeX macros.
\end{compactenum}

@@

The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
events~\cite{oopsla19a}. As the GNU R bytecode compiler is written in R, we turn
it off to avoid recording \eval in its code, and fallback on the interpreter.
This study requires recording a large number of events, calls to \eval,
side-effects, function entries and exits, environments, etc. One challenge we
encountered during the analysis was that R does not produce source references
for top-level blocks. That is to say, a method whose body would \c{eval(x)}
would not have debug information that we can use to identify that particular
\eval. The same expression surrounded by braces would have a line number and
source reference. This is unfortunately not easily fixed in the R
implementation. Instead, we extended the dynamic analysis tool to add missing
source references heuristically.

The analysis has some limitation. Even with the extension described above, there
are still \CranUndefinedRnd \eval calls without source references. This occurs
when a reference to the \eval function is passed as argument to a higher-order
functions or when the \eval call originates from native code. However, these
missing source references account for a meager \CranUndefinedRatio of all calls;
they are unlikely to affect our results. Other limitations are that we ignore
calls to the native \eval function and to the alternate \c{rlang::tidy\_eval}
function which uses native \eval internally. The \c{rlang} package introduces a
new kind of promise called \c{quosure} which is evaluated by \c{tidy\_eval}.
None of these limitation should invalidate our conclusions.
